---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# labtools

<!-- badges: start -->
<!-- badges: end -->

The goal of labtools is to provide a set of tools to help facilitate the 
handling of data that is commonly used in analytical chemistry laboratories,
for example extracting metadata for chemicals from Pubchem and then building and
viewing structure databases, exporting databases for MS-FINDER and MS-DIAL, 
compiling 2 dimensional GC QTOF/MS results exported from Canvas. It is still under
active development. In the future, more functions will be incorporated based on
the need of our lab.

## Release notes (2023.08.28 Version 0.0.3.0000)
1. Deposite functions to extract chemical metadata from Pubchem previouly incorporated 
in the fcmsafety package [GitHub](https://github.com/QizhiSu/fcmsafety) and update
them accordingly. Further updates will be done in this package while those in the
fcmsafety package will remain still.

## Installation

You can install the development version of labtools from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("QizhiSu/labtools")
```

## Read and combine Canvas data

To combine Canvas data, we first need to manually analyze GC x GC data in Canvas,
mark down peak of interest, and then export marked data in .txt format. All .txt 
files should be put into a folder as the function will read all .txt in the same
folder and then combine them into a single table by matching the chemical name. 
It will evaluate the retention index and second dimensional retention time of a 
same compound across them samples. If the differences are bigger than the defined 
tolerance, it will tell you which samples have significantly different RI or 2D RT,
such that you can carefully check where the inconsistencies locate. 
```{r read canvas data, eval = FALSE}
library(labtools)

data_path <- 'C:/data'
data <- read_canvas(data_path, 
                    ri_align_tolerance = 5,
                    rt_2d_tolerance = 0.05,
                    keep = 'area')

# to understand each argument, you can use the following code
?read_canvas

```

## Convert list of chemicals into structure database used by MS-FINDER
To convert a list of chemicals into structure database, we need to put the list 
of chemicals in either a txt, csv, xlsx, or xls file. Chemical name must be in the 
first column, and specify its path. If CAS is available, please specify in which 
column the CAS number locates. We encourage to have CAS number as it will be 
beneficial for extraction chemical information from Pubchem. Finally, we have to 
specify the path to store the structure database.
```{r export for msfinder, eval = FALSE}
library(labtools)

export4msfinder("c:/data/list_of_chemicals.xlsx",
                cas_col = 2,
                "c:/data/structure_database_for_msfinder.txt")

# to understand each argument, you can use the following code
?export4toxtree

```

